Generalized Measurement Invariance Tests for Factor Analysis

### 2012-01-24 www/psychoco2012/invariance.tex
Revised psychoco 2012 slides: - Added "overlay" figures to demonstrate invariance - Modified notation to match paper - Added some detail on software - Added some detail on test statistic construction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@89 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% TODO: Add something like Fig 3 from paper to show test stat construction?

@@

% FA path diagram, types of MI
% TODO: Overlays with different parameter values on graphic
\frame{
  \frametitle{Studying Measurement Invariance}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{famod.pdf}
  \end{center}}

% Define as theta_i = theta_0, i=1,..,n
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    % \item There exist many types of measurement invariance,
    %   based on specific subsets of parameters being allowed
    %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
    \item Hypothesis of ``full'' measurement invariance ($\bm{\theta}_i$ is the full parameter vector for individual $i$):\\
      \begin{center}
          $H_0: \bm{\theta}_i = \bm{\theta}_0, i=1,\ldots,n$\\
          $H_1: \text{Not all the }\bm{\theta}_i = \bm{\theta}_0$
      \end{center}
  \end{itemize}}

% Too difficult to assess, so use an auxiliary variable to place
% people in meaningful groups
  % continuous vs categorical auxiliary variable
\frame{
  \frametitle{Studying Measurement Invariance}
  \begin{itemize}
    \item $H_0$ from the previous slide is difficult to fully assess
      due to all the ways by which individuals may differ.\\ \ \\
    \item We typically place people into groups based on a meaningful
      auxiliary variable, then study measurement invariance across
      those groups (via Likelihood Ratio tests, Lagrange multiplier
      tests, Wald tests).
  \end{itemize}}

@@

% TODO Consistent notation with paper
\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item The proposed family of tests rely on first derivatives of
      the model's log-likelihood function.
    \item We can also consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}}) =
  {\bm 0},\text{ where} \\
 \\
  \psi(\bm{x}_i, \hat{{\bm \theta}}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

@@

\frame{
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by an auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
      \end{itemize}
  \end{itemize}
  \begin{center}
    $\bm{W}_n(t,\bm{\theta}) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).$\\ \ \\
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
\end{center}}

@@

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \begin{equation}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}}) 
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot),$
  \end{equation}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm W}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge (where $p$ is the number of parameters).
  \end{itemize}
}

@@

% TODO: Add some detail on software
\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we use
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item A combination of \texttt{lavaan} and our own code for calculating scores of fitted models.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
   % TODO: URL for working paper
  \end{itemize}}

>>>>

@@

% FA path diagram, types of MI
\frame{
  \frametitle{Studying Measurement Invariance: Age <= 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_under16.pdf}
  \end{center}}

\frame{
  \frametitle{Studying Measurement Invariance: Age > 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_over16.pdf}
  \end{center}}

% Define as theta_i = theta_0, i=1,..,n
\frame{
  \frametitle{Hypotheses}
  \begin{itemize}
    % \item There exist many types of measurement invariance,
    %   based on specific subsets of parameters being allowed
    %   to vary across individuals (e.g., Meredith, 1993).\\ \ \\
    \item Hypothesis of ``full'' measurement invariance ($\bm{\theta}_i$ is the full parameter vector for individual $i$):\\
      \begin{center}
          $H_0: \bm{\theta}_i = \bm{\theta}_0, i=1,\ldots,n$\\
          $H_1: \text{Not all the }\bm{\theta}_i = \bm{\theta}_0$
      \end{center}
  \end{itemize}}

% Too difficult to assess, so use an auxiliary variable to place
% people in meaningful groups
  % continuous vs categorical auxiliary variable
\frame{
  \frametitle{Hypotheses}
  \begin{itemize}
    \item $H_0$ from the previous slide is difficult to fully assess
      due to all the ways by which individuals may differ.\\ \ \\
    \item We typically place people into groups based on a meaningful
      auxiliary variable, then study measurement invariance across
      those groups (via Likelihood Ratio tests, Lagrange multiplier
      tests, Wald tests).
  \end{itemize}}

\frame{
  \frametitle{Lack of Grouping}
  \begin{itemize}
    \item If we did not know the groups in advance, we could conduct a LRT wrt to all possible groups, taking the maximum.  The proposed tests are related to this idea.
  \end{itemize}
 % TODO Insert Fig 3
}

@@

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item The proposed family of tests rely on first derivatives of
      the model's log-likelihood function.
    \item We can also consider individual terms ({\em scores}) of the gradient.  These scores tell us how well a particular parameter describes a particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
  {\bm 0},\text{ where} \\
 \\
  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

@@

\frame{
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by an auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
      \end{itemize}
  \end{itemize}
  \begin{center}
    $\bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).$\\ \ \\
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
\end{center}}

@@

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \begin{equation}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),$
  \end{equation}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm B}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge ($p$ is the number of parameters).
  \end{itemize}
}

@@

\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we utilize
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item \texttt{estfun()} for score extraction, which is (currently) a combination of our own code and \texttt{lavaan} code.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
          \begin{itemize}
            \item Required input: Fitted model, function for score extraction, and information matrix (optional)
            \item \texttt{gefp()} constructs the process.
            \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
          \end{itemize}
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
    \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}
  \end{itemize}}
